Ricci-Ollivier curvature of two random walks on rooted phylogenetic subtree-prune-regraft graph

### 2015-03-10 curvature.tex
a few tweaks in response to ec79c6dcdf72aa0f15c582a6eb49e3c9be33734b

\subsection{Computing rSPR treespaces}

@@

\begin{theorem}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	We implement the graph as an adjacency list with integer-labelled vertices.
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$).
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick representation of the tree.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes and pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to realistic phylogenetic trees).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}
This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}, which outputs an edge list format graph suitable for input to other software. \todo{github url}
Our fast graph construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm requires only 303.45 seconds to construct it.
Constructing the 8-taxon tree space is not feasible for the previous method in a reasonable amount of time.
Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.

\subsection{Computing curvature values}
We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
We first used \todo{\dots} to compute the rSPR treespaces for four to seven leaves.
This required \dots time.
We then used linear programming to compute curvatures for given pairs of trees \cite{gricci}.
This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} (enumeration methods described in a manuscript in preparation).

@@

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves such that an arbitrary SPR neighbor as described in the next paragraph.
Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.

>>>>

\subsection{Computing rSPR treespaces}
\label{sec:computing_treespace}

@@

\begin{theorem}
	The subgraph of the rSPR graph induced by a set $\mathcal{T}$ of $m$ trees with $n$ leaves can be constructed in $\OhOf{mn^3}$-time.
\end{theorem}
\begin{proof}
	We implement the graph as an adjacency list with integer-labelled vertices.
	As described above, the integer labels are simply the order of the input trees.
	Adding the vertices to the graph requires $\OhOf{m}$-time, as they are added in ascending order to the end of the vertex list.
	Adding the $\OhOf{mn^2}$ edges to the graph requires $\OhOf{mn^2 \log n}$-time (as the vertex degrees are $\OhOf{n^2}$).
	Enumerating the neighbors of $T$ requires $\OhOf{n^3}$-time for each $T$, for a total of $\OhOf{mn^3}$-time.
	We discuss below, in Section~\ref{sec:random_walks} how to do so efficiently without considering duplicate neighbors.
	We store the tree to index mappings for current vertices of $G$ in a trie using the $\OhOf{n}$-length Newick representation of the tree.
	These representations are made unique by ordering the tree so that leftmost subtrees contain the smallest alphanumeric label of descendants.
	This requires only $\OhOf{n}$-time for each tree (i.e. a total of $\OhOf{mn^3}$-time) using a standard nodes and pointers representation of the tree and assuming integer leaf labels (a simple $\OhOf{mn\log n}$ leaf preprocessing step could be applied to extend this procedure to realistic phylogenetic trees).
	Similarly, it takes $\OhOf{n}$-time to determine the index of each of the $\OhOf{mn^2}$ $S$'s.
	Therefore the graph can be constructed in $\OhOf{mn^3}$-time.
\end{proof}
This procedure has been implemented in the C++ program \texttt{dense\_spr\_graph} of the software package \texttt{spr\_neighbors}, which outputs an edge list format graph suitable for input to other software. \todo{github url}
Our fast graph construction procedure reduced the time required to compute the 10,395-vertex 7-taxon rSPR tree space from 2,104.68 seconds to 12.71 seconds on an Intel Core 2 Duo E7500 desktop running Ubuntu 14.04.
Moreover, although we do not study the 135,135-vertex 8-taxon tree space in this paper, our algorithm requires only 303.45 seconds to construct it.
Constructing the 8-taxon tree space is not feasible using the previous method in a reasonable amount of time. \todo{Going on 92 hours at the moment. I'll give it a bit longer and then just say that we terminated our attempt after X hours}
Thus, we believe this fast graph construction procedure will itself be useful for further studies of treespace subsets similar to~\cite{Whidden2015-yi}.

\subsection{Computing curvature values}
We computed all of the curvatures for pairs of trees with up to seven leaves to gain an intuitive understanding of the distribution of curvatures.
We first used \texttt{dense\_spr\_graph} to compute the rSPR treespaces for four to seven leaves, as discussed in Section~\ref{sec:computing_treespace}.
We then used linear programming to compute curvatures for given pairs of trees \cite{gricci}.
This would have required an enormous amount of computation to directly compute curvatures for the $((2n-3)!!)^2$ pairs of trees with $n$ leaves.
We instead exploited the fact that pairs of trees which are equivalent modulo label renumbering are symmetric in the rSPR graph and therefore guaranteed to have the same curvature.
We thus directly computed curvature values for one representative pair from each such equivalence class, or \emph{tanglegram} (enumeration methods described in a manuscript in preparation).

@@

To efficiently simulate the MH random walk, we developed a linear-time algorithm for proposing SPR moves that does not require the rSPR graph to be explicitly built.
A na\"ive approach would require $\OhOf{n^3}$ time---$OhOf{n}$ time to generate each of the $\OhOf{n^2}$ neighbors of a given tree--- and then one could be picked uniformly at random.
To eliminate an $\OhOf{n^2}$ factor, we developed a deterministic ordering of SPR moves with a one-to-one correspondence to SPR neighbors, as described in the next paragraph.
Given such an order, a uniform neighbor can be selected by its index in $\OhOf{n}$ time.
We note that the recursive formula of Song \cite{Song2003-gf} for the degree of a tree does not group SPR moves that move a particular subtree, and thus would still require $\OhOf{n^2}$ time to select a specific SPR neighbor by index.
